Longitudinal Data Cleaning

### 2021-02-24 report_alt.Rmd
Edit the report template and add rlm section.

---
title: Longitudinal Data Cleaning - A Case of the NLSY79 Data
authors:
  - name: Dewi Amaliah
    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Department of Econometric and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3168
    email: dama0007@student.monash.edu
  - name: Dianne Cook
    department: Department of Econometric and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3168
    email: dicook@monash.edu
abstract: |
  Enter the text of your abstract here.
keywords:
  - longitudinal data
  - data cleaning 
  - data tidying
  - robust linear model
bibliography: references.bib
biblio-style: unsrt
output: 
  rticles::arxiv_article:
    keep_tex: true
---

@@

@@

@@

```{r demoprofile, echo = FALSE, fig.cap="Demography Profile of the NLSY79 Cohort", fig.width=6}
age <- yowie::demographic_nlsy79 %>%
  ggplot(aes(x = age_1979)) +
  geom_bar(fill = "#45B39D") +
  ggtitle("age distribution") +
  theme(plot.title = element_text(size = 10))

race <- yowie::demographic_nlsy79 %>%
  ggplot(aes(x = race,
             fill = race)) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count", vjust = 1.5, colour = "white") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 10)) +
  ggtitle("race_distribution") +
  theme(plot.title = element_text(size = 10),
        axis.text.x = element_text(size = 6))

@@

```{r spaghetti, echo = FALSE}
spag <- wages_demog_hs %>%
  ggplot(aes(x = year,
             y = mean_hourly_wage,
             group = id)) +
  geom_line(alpha = 0.1) +
  ggtitle("a)") +
  theme(plot.title = element_text(size = 10))
```

```{r featureplot, echo = FALSE, fig.cap = "a. The mean hourly wage of cohort whose highest grade completed is up to 12th grade, b. Three features of the cohort", fig.width=6, fig.height=4}
wages_hs2020 <- as_tsibble(x = wages_demog_hs,
                    key = id,
                    index = year,
                    regular = FALSE)
wages_three_feat <- wages_hs2020 %>%
  features(mean_hourly_wage, 
           feat_three_num
           )
wages_feat_long <- wages_three_feat %>%
  pivot_longer(c(min, med, max), names_to = "feature", values_to = "value")
 feature <- ggplot(wages_feat_long) +
  geom_density(aes(x = value, colour = feature, fill = feature), alpha = 0.3) +
   ggtitle("b)") +
   theme(plot.title = element_text(size = 10))

spag + feature
```
```{r summarytable, echo = FALSE}
kable(as.array(summary(wages_demog_hs$mean_hourly_wage)), 
      caption = "Summary Statistics of Wages of High School Data",
      col.names = c("Statistics", "Value")) %>%
  kable_styling(latex_options = "striped")
```

Further, we took 36 samples randomly from the data and plotted it as is seen in Figure \ref{fig:sampleplot}. It implies that not only that some observations earned an extremely high figures of wages, but some also had a very fluctuative wages, fo example in panel number 5, 7, and 11.

```{r sampleplot, echo = FALSE, fig.cap="Some samples of observations"}
set.seed(20210225)

ggplot(wages_hs2020, 
       aes(x = year,
                y = mean_hourly_wage,
                group = id)) +
  geom_line(alpha = 0.7) +
  facet_sample() +
  ylab("mean hourly wage") 
```

### Robust Linear Model for Influential Values Treatment

@@

>>>>

---
title: "Longitudinal Data Cleaning"
subtitle: "A Case of The NLSY79 Data"
author: "Dewi Lestari Amaliah"
date: "24/02/2021"
output: 
  bookdown::pdf_document2:
    toc : false
bibliography: references.bib
---

@@

Real world or wild caugh data

@@

```{r demoprofile, echo = FALSE, fig.cap="Demography Profile of the NLSY79 Cohort", fig.width=5}
age <- yowie::demographic_nlsy79 %>%
  ggplot(aes(x = age_1979)) +
  geom_bar(fill = "#45B39D") +
  ggtitle("age distribution") +
  theme(plot.title = element_text(size = 10))

race <- yowie::demographic_nlsy79 %>%
  ggplot(aes(x = race,
             fill = race)) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count", vjust = 1.5, colour = "white") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 10)) +
  ggtitle("race_distribution") +
  theme(plot.title = element_text(size = 10),
        axis.text.x = element_text(size = 4))

@@

```{r featureplot, echo = FALSE, fig.cap = "a) Pattern of Mean Hourly Wage of High School Cohort; b) The Three Features of the Cohort", fig.width=6, fig.height=4}

spag <- wages_demog_hs %>%
  ggplot(aes(x = year,
             y = mean_hourly_wage,
             group = id)) +
  geom_line(alpha = 0.1) +
  ggtitle("a)") +
  theme(plot.title = element_text(size = 10))

wages_demog_hs_tsibble <- as_tsibble(x = wages_demog_hs,
                    key = id,
                    index = year,
                    regular = FALSE)
wages_three_feat <- wages_demog_hs_tsibble %>%
  features(mean_hourly_wage, 
           feat_three_num
           )
wages_feat_long <- wages_three_feat %>%
  pivot_longer(c(min, med, max), names_to = "feature", values_to = "value")
 feature <- ggplot(wages_feat_long) +
  geom_density(aes(x = value, colour = feature, fill = feature), alpha = 0.3) +
   ggtitle("b)") +
   theme(plot.title = element_text(size = 10))

spag + feature
```

@@

```{r summarytable, echo = FALSE}
kable(as.array(summary(wages_demog_hs$mean_hourly_wage)), 
      caption = "Summary Statistics of Wages of High School Data",
      col.names = c("Statistics", "Value")) %>%
  kable_styling(latex_options = "striped")
```

Further, we took 36 samples randomly from the data and plotted it as is seen in Figure \ref{fig:sampleplot}. It implies that not only that some observations earned an extremely high figures of wages, but some also had a very fluctuate wages, fo example in panel number 5, 7, and 11.

```{r sampleplot, echo = FALSE, fig.cap="Some samples of observations", fig.height=4, fig.width=5}
set.seed(20210225)

ggplot(wages_demog_hs_tsibble, 
       aes(x = year,
                y = mean_hourly_wage,
                group = id)) +
  geom_line(alpha = 0.7) +
  facet_sample() +
  theme(axis.text.x = element_text(angle = 10, size = 6)) +
  ylab("mean hourly wage") 
```

### Robust Linear Model for Noises Treatment

As it is seen from figure \ref{fig:sampleplot}, there are many spikes in the mean hourly wage data. As part of the IDA, which is the model formulation, we built a robust linear regression model to address this issue. The notion of robust linear regression is to yield an estimation that is robust to the influence of noise or contamination [@KollerManuel2016rARP]. It also aims to detect the contamination by weighting each observation based on how "well-behave" they are, known as robustness weight. Observations with lower robustness weight are suggested as an outliers by this method [@KollerManuel2016rARP].

In this paper, we built the model using the `rlm` function from `MASS` package [@mass]. We set the `mean_hourly_wage` and `year` as the dependent and predictor respectively. Furthermore, we used M-Estimation with Huber weighting where the observation with small residual get a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm].

Since we worked with longitudinal data, we should built the model for each ID, instead of the overall data. The robust mixed model is actually the best model to be employed in this case. However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data. Thus, the model for each ID is built utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr] respectively. After getting the weight values, we flagged the observations with the weight less than 1, and imputed their mean hourly wage with the models' predicted value.

Figure \ref{fig:comppict} a) shows that after imputing the outliers with the models' predicted value, the highest wages value is not exceed US$200. The spikes were still observed, but are not as extreme as the original data. In Figure \ref{fig:comppict} b), although the distributions of the features are still positively skewed, we can still examine it clearly. We also learned that some IDs' have a minimum wages that is higher than others' maximum wages.

```{r rlm, cache=TRUE, warning = FALSE, message = FALSE}

# nest the data by id to build a robust linear model
by_id <- wages_demog_hs %>%
  dplyr::select(id, year, mean_hourly_wage) %>%
  group_by(id) %>%
  nest()

# build a robust linear model
id_rlm <- by_id %>%
  mutate(model = map(.x = data,
                     .f = function(x){
                       rlm(mean_hourly_wage ~ year, data = x)
                     }))
# extract the property of the regression model
id_aug <- id_rlm %>%
  mutate(augmented = map(model, broom::augment)) %>%
  unnest(augmented)

# extract the weight of each observation
id_w <- id_rlm %>%
  mutate(w = map(.x = model,
                 .f = function(x){
                   x$w
                 })) %>%
  unnest(w) %>%
  dplyr::select(w)

# bind the property of each observation with their weight
id_aug_w <- cbind(id_aug, id_w) %>%
  dplyr::select(`id...1`,
                year,
                mean_hourly_wage,
                .fitted,
                .resid,
                .hat,
                .sigma,
                w) %>%
  rename(id = `id...1`)

# if the weight < 1, the mean_hourly_wage is replaced by the model's fitted/predicted value.
# and add the flag whether the observation is predicted value or not.
# since the fitted value is sometimes <0, and wages value could never be negative,
# we keep the mean hourly wage value even its weight < 1.

wages_rlm_dat <- id_aug_w %>%
  mutate(wages_rlm = ifelse(w != 1 & .fitted >= 0, .fitted,
                            mean_hourly_wage)) %>%
  mutate(is_pred = ifelse(w != 1 & .fitted >= 0, TRUE,
                          FALSE)) %>%
  dplyr::select(id, year, wages_rlm, is_pred)

# join back the `wages_rlm_dat` to `wages_demog_hs`

wages_demog_hs <- left_join(wages_demog_hs, wages_rlm_dat, by = c("id", "year"))

```

```{r comppict, echo = FALSE, fig.cap="a) The Wages and b) The Feature Distribution after the Noises Treatment", fig.width=6, fig.height=4}

spag2 <- wages_demog_hs %>%
  ggplot(aes(x = year,
             y = wages_rlm,
             group = id)) +
  geom_line(alpha = 0.1) +
  ggtitle("a)") +
  theme(plot.title = element_text(size = 10))

wages_hs2020_rlm <- as_tsibble(x = wages_demog_hs,
                    key = id,
                    index = year,
                    regular = FALSE)
wages_three_feat_rlm <- wages_hs2020_rlm %>%
  features(wages_rlm, 
           feat_three_num
           )
wages_feat_long_rlm <- wages_three_feat_rlm %>%
  pivot_longer(c(min, med, max), names_to = "feature", values_to = "value")
feature2 <- ggplot(wages_feat_long_rlm) +
  geom_density(aes(x = value, colour = feature, fill = feature), alpha = 0.3) +
  ggtitle("b)") +
  theme(plot.title = element_text(size = 10))

spag2 + feature2
```

Furthermore, the robust linear regression has reduce the level of noise in the data set as is seen in Figure \ref{fig:compare}. The result is that we got a data set with the reasonable degree of fluctuation.

```{r compare, echo = FALSE, fig.cap="Comparison between the Original and Treated Mean Hourly Wage", fig.height=4, fig.width=5}
set.seed(20210228)

sample_id <- sample(unique(wages_demog_hs$id), 20)
sample <- subset(wages_demog_hs, id %in% sample_id)

wages_compare <- sample %>%
  dplyr::select(id, year, mean_hourly_wage, wages_rlm) %>%
  rename(mean_hourly_wage_rlm = wages_rlm) %>%
  pivot_longer(c(-id, -year), names_to = "type", values_to = "wages")

ggplot(wages_compare) +
  geom_line(aes(x = year,
                y = wages,
                colour = type,
                linetype = type),
            alpha = 1) +
  theme(axis.text.x = element_text(angle = 10, size = 5),
        legend.position = "bottom") +
  facet_wrap(~id)

```

Finally, we saved the imputed data and set the appropriate data type for the variables. We also saved the NLSY79 cohort's demographic information and the high school dropout cohort in a separate data sets.

```{r}
# select out the old value of mean hourly wage and change it with the wages_rlm value
wages_demog_hs <- wages_demog_hs %>%
  dplyr::select(-mean_hourly_wage) %>%
  rename(mean_hourly_wage = wages_rlm)

# rename and select the wages in tidy
wages_hs2020 <- wages_demog_hs %>%
  dplyr::select(id, year, mean_hourly_wage, age_1979, gender, race, hgc, hgc_i, yr_hgc,
                number_of_jobs, total_hours, is_wm, is_pred) %>%
  mutate(hgc = as.factor(hgc),
         year = as.integer(year),
         age_1979 = as.integer(age_1979),
         yr_hgc = as.integer(yr_hgc),
         number_of_jobs = as.integer(number_of_jobs))

# Create a data set for demographic variables
demographic_nlsy79 <- full_demographics %>%
  mutate(age_1979 = 1979 - (dob_year + 1900)) %>%
  dplyr::select(id,
         age_1979,
         gender,
         race,
         hgc,
         hgc_i,
         yr_hgc) %>%
  mutate(age_1979 = as.integer(age_1979),
         hgc = as.factor(hgc),
         yr_hgc = as.integer(yr_hgc))

# Create a data set for the high school dropouts cohort
wages_hs_dropout <- wages_hs2020 %>%
  mutate(dob = 1979 - age_1979,
         age_hgc = yr_hgc - dob) %>%
  filter((hgc %in% c("9TH GRADE",
                     "10TH GRADE",
                     "11TH GRADE")) |
           (hgc == "12TH GRADE" &
              age_hgc >= 19)) %>%
  dplyr::select(-dob,
         -age_hgc)

```

@@

# Reference
